Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-06 paper/ms.tex

Another special case of the \genrp\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is a normalization constant.
The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}

These identities yield a kernel of the form
\begin{equation}
k(\tau) = S_0 \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \begin{cases}
\cosh{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sinh{(\eta \omega_0 \tau)}, & 0 < Q < 1/2\\
2 (1+\omega_0 \tau), & Q = 1/2\\
\cos{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sin{(\eta \omega_0 \tau)},& 1/2 < Q < \infty\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4Q^2)^{-1}\vert^{1/2}$.
It is interesting to note that, because of the damping, the characteristic
oscillation frequency in this model $d_j$, for any finite quality factor $Q > 1/2$,
is not equal to the frequency of the undamped oscillator $\omega_0$.

@@

{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
\begin{eqnarray}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
\end{eqnarray}
A model like this has been used to model the granulation is Solar data
\citep{Michel:2009}}, which has a kernel of
\begin{equation}
k(\tau) = S_0 \omega_0e^{-\frac{1}{\sqrt{2}}\omega_0\tau} \cos{\left(\frac{\omega_0\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
\end{equation}

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $\omega_0/c_j = 2\,Q = 20$ and normalized so that
    $S(d_j)/S(0) = Q^2 = 100$.
    (right) The corresponding autocorrelation functions with the same colors.
    \figurelabel{sho}}
\end{center}
\end{figure}

>>>>

Another special case of the \genrp\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0$.
The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
    b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}

These identities yield a kernel of the form
\begin{equation}
k(\tau) = S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
\begin{cases}
    \cosh{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
    2\,(1+\omega_0\,\tau), & Q = 1/2\\
    \cos{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
It is interesting to note that, because of the damping, the characteristic
oscillation frequency in this model $d_j$, for any finite quality factor $Q > 1/2$,
is not equal to the frequency of the undamped oscillator $\omega_0$.

@@

{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
\begin{eqnarray}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
\end{eqnarray}
A model like this has been used to model the granulation is Solar data
\citep{Michel:2009}}, which has a kernel of
\begin{equation}
k(\tau) = S_0\,\omega_0\,e^{-\frac{1}{\sqrt{2}}\,\omega_0\,\tau}\,
    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
\end{equation}

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = \omega_0/2\,Q = 1/20$ and normalized so that
    $S(d_j)/S(0) = 100$.
    (right) The corresponding autocorrelation functions with the same colors.
    \figurelabel{sho}}
\end{center}
\end{figure}
